---
title: "Weight loss analysis"
author: "Md. Fardin Khan"
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: cerulean
    toc: true
    toc_float: true
    code_folding: show
---

# **Purpose of the Study**
Obesity and overweight conditions are major public health concerns. This study aims to analyze the **effectiveness of a 12-week weight loss program** using a dataset collected from 35 participants.I will examine the relationships or impact between **diet adherence, exercise intensity, sleep quality**, and **weight loss** using statistical methods in R.

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, warning = FALSE, message = FALSE)
```

# **Load Required Libraries**
```{r}
# Load necessary libraries
library(ggplot2)  
library(dplyr)     
library(corrplot)
library(readxl)
library(naniar)
library(car)
library(gtsummary)
```

# ** Import Dataset **
```{r}
data <- read_excel("Data_WeightLoss.xlsx")
head(data)
```
# ** EDA **
```{r}
isnull <- colSums(is.na(data))
print(paste("Missing value count:", isnull))

gg_miss_var(data,show_pct = TRUE)+
  labs(y = "Displaying the proportion of missings")
```

```{r}
numeric_cols <- data[sapply(data, is.numeric)]
stats <- data.frame(
  Mean = sapply(numeric_cols, mean, na.rm = TRUE),
  Std = sapply(numeric_cols, sd, na.rm = TRUE),
  Min = sapply(numeric_cols, min, na.rm = TRUE),
  Q1 = sapply(numeric_cols, function(x) quantile(x, 0.25, na.rm = TRUE)),
  Median = sapply(numeric_cols, median, na.rm = TRUE),
  Q3 = sapply(numeric_cols, function(x) quantile(x, 0.75, na.rm = TRUE)),
  Max = sapply(numeric_cols, max, na.rm = TRUE),
  IQR = sapply(numeric_cols, IQR, na.rm = TRUE)
)

stats$Lower_Bound = stats$Q1 - 1.5*stats$IQR
stats$Upper_Bound = stats$Q3 + 1.5*stats$IQR

print(stats)

```
## Data Inspection and Observations

Upon inspecting the dataset, the following observations were made:

- **Missing values**: The dataset has **1 missing value** in the `ExerciseIntensity` column.
- **Gender column**: The `Gender` column is in **character format**. It may be beneficial to convert this column into a numeric format for further analysis.
- **Participant column**: The `Participant` column appears to be a **serial number** rather than a categorical or numeric variable of interest.

- **Outlier**: As per the `Lower_Bound` and `Upper_Bound` of the each column shows that there is no outlier value


```{r}
# Drop Participant col
data <- data %>% select(-Participant)
# Convert Gender to factor
data$Gender <- as.factor(data$Gender)
head(data)
```

```{r fig.width=10, fig.height=5}
# checking up the correlation with columns and convert the gender factor to numeric and store in "Gender_numeric" column
data$Gender_numeric <- as.numeric(factor(data$Gender))

cor_matrix <- cor(data[, c("WeightLoss", "BaselineWeight","Age", "AfterWeight", "DietAdherence", "ExerciseIntensity", "SleepQuality", "Gender_numeric")], use="complete.obs")

corrplot(cor_matrix, method="number", type="full",order = 'AOE',col= COL2('PuOr', 40),)

print(cor_matrix)

```

## Interpretation of Correlation Results

From the correlation matrix, we observe that the `WeightLoss` attribute has significant correlations with several other numeric variables. Specifically:

- **Positive correlations**: 
   - `DietAdherence` and `ExerciseIntensity` show strong positive correlations with `WeightLoss`.
   - `SleepQuality` and `Gender_numeric` also exhibit positive correlations with `WeightLoss`.

- **Negative correlations**: 
   - `BaselineWeight`, `AfterWeight` and `Age` have negative correlations with `WeightLoss`, indicating that as these variables increase, weight loss tends to decrease.

Overall, we can conclude that `DietAdherence` and `ExerciseIntensity` are key factors positively influencing weight loss, while `BaselineWeight` and `Age` are negatively related.

## Data Visualization
```{r}
for (i in names(data)){
  if (i %in% c("WeightLoss","BaselineWeight","Age", "DietAdherence", "ExerciseIntensity", "SleepQuality")){
   print(
     ggplot(data, aes_string(x = i)) +
        geom_histogram(color = "black", bins = 5, alpha = 0.6)  +
        labs(title = paste("Histogram of ", i), x = i, y = "Frequency") +
        theme_minimal()
   )
  }
}

```

# ** Regression with threshold = 0.3 **
In here if i select `threshold = 0.3` than columns will:

- **DietAdherence**
- **ExerciseIntensity**
- **Age**
- **AfterWeight**
```{r}
model <- lm(WeightLoss ~ Age + AfterWeight + DietAdherence + ExerciseIntensity, data = data)

model %>%
  tbl_regression() %>%
  bold_labels() %>%
  bold_p(t=.1)

```
```{r}
summary(model)
```

```{r}
qqnorm(resid(model))
qqline(resid(model))

```

```{r}
#plot(model$fitted.values, resid(model))
#abline(h=0, col="red")

```

```{r}
print(vif(model))

barplot(vif(model), main = "VIF Plot", col = "lightblue", ylim = c(0, max(vif(model)) + 10))
abline(h = 10, col = "red", lty = 2)

```

## Finding on threshold = 0.3
My Model demonstrates exceptional statistical significance with an `F-statistic of 26.03` and a ` p-value of 3.117e-09`.With an `R-squared value of 0.78` indicates that our model explains `78%` of the variance in weight loss outcomes. [[4](https://statisticsbyjim.com/regression/interpret-r-squared-regression/#:~:text=R%2Dsquared%20is%20a%20goodness%2Dof%2Dfit,variance%20in%20the%20dependent)][[5](https://en.wikipedia.org/wiki/P-value)]

In here the residuals range from `-1.62` to `1.38`, with a median close to zero, which is generally a good sign.

All `Variance Inflation Factor (VIF)` values in our model are below `1.06`, this ensure that model coefficient estimates are stable and that model can confidently interpret the individual effects of each predictor on weight loss outcomes.
[[1](https://www.numberanalytics.com/blog/essential-insights-vif-regression#:~:text=The%20Variance%20Inflation%20Factor,the%20predictor%20variables.%20In)][[2](https://pmc.ncbi.nlm.nih.gov/articles/PMC6900425/#:~:text=multicollinearity%2C%20measures%20to%20detect,text%2C%20the%20concepts%20of)][[3](https://www.analyticsvidhya.com/blog/2020/03/what-is-multicollinearity/)]

The Q-Q plot of residuals demonstrates a normal distribution 
, further validating the model's assumptions and strengthening the credibility of our findings. This normal distribution of residuals indicates that our model is well-specified and that the relationships between variables are appropriately captured.[[6](https://en.wikipedia.org/wiki/Q%E2%80%93Q_plot)]

| **Characteristic**     |  **Interpretation**                                           |
|------------------------|--------------------------------------------------------------|
| **Age**                | As age increases, weight loss decreases slightly.             |
| **AfterWeight**        | Higher after-weight slightly reduces weight loss.             |
| **DietAdherence**      | Higher diet adherence significantly increases weight loss.    |
| **ExerciseIntensity**  | Higher exercise intensity significantly increases weight loss.|


# ** Regression with threshold = 0.2 **
In here if i select `threshold = 0.2` than columns will:

- **DietAdherence**
- **ExerciseIntensity**
- **Age**
- **AfterWeight**
- **SleepQuality**
- **Gender_numeric**
```{r}
model2 <- lm(WeightLoss ~ Age + AfterWeight + DietAdherence + ExerciseIntensity +SleepQuality+ Gender_numeric, data = data)

model2 %>%
  tbl_regression() %>%
  bold_labels() %>%
  bold_p(t=.1)

```


```{r}
summary(model2)
```

```{r}
qqnorm(resid(model2))
qqline(resid(model2))

```


```{r}
print(vif(model2))

barplot(vif(model2), main = "VIF Plot", col = "lightblue", ylim = c(0, max(vif(model2)) + 10),las = 2,      cex.names = 0.7)
abline(h = 10, col = "red", lty = 2)
```


## Finding on threshold = 0.2
My Model demonstrates exceptional statistical significance with an `F-statistic of 63.32  and a `p-value: 1.231e-14`.With an `R-squared value of 0.93` indicates that our model explains `93%` of the variance in weight loss outcomes [[4](https://statisticsbyjim.com/regression/interpret-r-squared-regression/#:~:text=R%2Dsquared%20is%20a%20goodness%2Dof%2Dfit,variance%20in%20the%20dependent)].The highly significant F-statistic (p < 0.001) provides robust evidence for the model's overall validity.[[5](https://en.wikipedia.org/wiki/P-value)]

In here the residuals range from `-0.81299` to `0.81801`, with a median close to zero, which is generally a good sign.

All `Variance Inflation Factor (VIF)` values in our model are below `1.12` and below the threshold, this ensure that model coefficient estimates are stable and that model can confidently interpret the individual effects of each predictor on weight loss outcomes.
[[1](https://www.numberanalytics.com/blog/essential-insights-vif-regression#:~:text=The%20Variance%20Inflation%20Factor,the%20predictor%20variables.%20In)][[2](https://pmc.ncbi.nlm.nih.gov/articles/PMC6900425/#:~:text=multicollinearity%2C%20measures%20to%20detect,text%2C%20the%20concepts%20of)][[3](https://www.analyticsvidhya.com/blog/2020/03/what-is-multicollinearity/)]

The Q-Q plot of residuals demonstrates a normal distribution 
, further validating the model's assumptions and strengthening the credibility of our findings. This normal distribution of residuals indicates that our model is well-specified and that the relationships between variables are appropriately captured.[[6](https://en.wikipedia.org/wiki/Q%E2%80%93Q_plot)]

| **Characteristic**     |  **Interpretation**                                           |
|------------------------|--------------------------------------------------------------|
| **Age**                | As age increases, weight loss decreases slightly.             |
| **AfterWeight**        | Higher after-weight slightly reduces weight loss.             |
| **Gender_numeric**     | Indicates potential differences in weight loss outcomes between genders.            |
| **SleepQuality**       | Moderate positive effect             |
| **DietAdherence**      | Second strongest positive effect    |
| **ExerciseIntensity**  | Strongest positive predictor of weight loss.|


# **Final Finding**

### **Model Comparison and Key Findings**

| Predictor        | Coefficient (Threshold 0.3) | Coefficient (Threshold 0.2) | p-value (Threshold 0.2)  | p-value (Threshold 0.3)  | Interpretation                                    |
|------------------|----------------------------|----------------------------|-------------------------|-------------------------|--------------------------------------------------|
| **Age**          | -0.04238                   | -0.030569                  | 0.00144                 | 0.007                 | Older age decreases weight loss.                |
| **AfterWeight**  | -0.02999                   | -0.025912                  | 0.00393                 | 0.044                 | Higher starting weight decreases weight loss.   |
| **DietAdherence**| 0.60710                    | 0.624733                   | <0.001                  | <0.001                  | Higher diet adherence increases weight loss.    |
| **ExerciseIntensity** | 0.64168               | 0.728287                   | <0.001                  | <0.001                  | Higher exercise intensity increases weight loss.|
| **SleepQuality** | N/A                        | 0.404399                   | <0.001                  | N/A                  | Better sleep quality increases weight loss.     |
| **Gender_numeric**| N/A                       | 0.388966                   | 0.00712                 | N/A                  | Gender positively impacts weight loss.          |

### **VIF (Variance Inflation Factor)**:
All predictors have low VIF values (< 1.2), indicating no significant multicollinearity.

### **Statistical Significance**:
- **DietAdherence** and **ExerciseIntensity** are highly significant in both models.
- **Age** and **AfterWeight** are significant predictors, with negative effects on weight loss.
- **SleepQuality** and **Gender** show positive effects in the extended model.

### **Practical Implications**:
- Focus on improving **DietAdherence** and **ExerciseIntensity** for better weight loss outcomes.
- Incorporate **SleepQuality** as part of the program for additional benefits.
- **Age** and **AfterWeight** should be considered for personalized interventions.

### **Limitation**:
- **Sample Size:** With only 35 participants, the generalizability of the findings might be limited. Larger sample sizes are needed for more robust conclusions.
- **External Factors:** The analysis only considers a few lifestyle factors, but other factors such as psychological health, stress, and external environmental influences were not accounted for.

### **Conclusion**:
Both models emphasize the importance of **DietAdherence** and **ExerciseIntensity** for weight loss. The extended model with **SleepQuality** and **Gender** provides a more comprehensive understanding. Tailored programs based on these factors will be more effective in helping individuals achieve weight loss goals.